Type Ia Supernova Explosions from Hybrid Carbon-Oxygen-Neon White Dwarf Progenitors That Have Mixed During Cooling

### 2019-06-26 paper/ms.tex
Went though the paper and fixed a good few things. Picked off some of the questions in blue or red text and in so doing added a little text and citations. Updated the references and fixed a few typos.

Thermonuclear (Type Ia) supernovae (\SNeIa) are bright stellar explosions
thought to occur when approximately 1.0 \Msun\ of material composed principally
of C and O burns under degenerate conditions. This class of supernovae is
known to synthesize much of the Fe-group elements found in the galaxy, and
the light curves of these events have a special property that allows
standardization \blust{of their light curves and thus use}
as distance indicators for cosmological studies~\citep{phillips:absolute}.
This use resulted in the discovery of the acceleration of the expansion of
the Universe and thus the inference of Dark
Energy~\citep{riess.filippenko.ea:observational,
perlmutter.aldering.ea:measurements,leibundgut2001}, and these events
remain the premier distance indicators for cosmological studies {\color{red} REF?}.
The special property of the light curve is thought to follow
from the fact that the source of luminosity, the radioactive decay
of \Ni{56}, synthesized by the thermonuclear burning, is also the
principal source of opacity~\citep{Pinto2001The-type-Ia-sup}, giving
the Phillips relation between the brightness of an event and the
rate of decline of the B-band magnitude from maximum~\citep{phillips:absolute}.

@@

The single degenerate picture posits a white dwarf gaining mass
from a main-sequence companion. The process relies on a long
period of accretion combined with either steady burning or a
series of nova explosions. {\color{blue} During this accretion phase, the WD must} gain at least
$\sim 0.4$ \Msun\ as needed for it to approach the
limiting Chandrasekhar mass~\citep{starrfieldetal2012}. As it approaches
the Chandrasekhar limit, conditions in the compressed core are right
to ignite the thermonuclear burning that will incinerate the star.

@@

A particular problem was that most models included a massive He layer on
the white dwarf. Such a configuration leads to the unobserved result that
intermediate and heavy elements synthesized in the He detonation appear in
the outer parts of the ejecta, which does not match observation~\citep{HoefKhok96,
hoeflichetal96,finkhillebrandtroepke2007,simetal2010}.  \citet{bildstenetal2007},
however, found that fairly thin He layers could flash on sub-Chandrasekhar
mass white dwarfs, partially resolving the issue and encouraging
further research~\citep{simetal2012,brooksetal2015, shenetal2018,
glasneretal2018}.

@@

Subsequent work allayed this concern {\color{blue} [we should probably mention how]}
~\citep{yoonetal2007,lorenaguilaretal2009, Shenetal12, pakmoretal2012b}
and active research continues, motivated by the need to explain bright events
and the estimated population of WD pairs. Contemporary
research focuses on variations on the merger idea, including inspiraling pairs,
collisions, violent mergers, and the ``core-degenerate'' model in which the merger
takes place in a common envelope~\citep{raskinetal2009,pakmoretal2011,kashi:2011,
pakmoretal2012a,Shenetal12,katzetal2016}.

@@

Modern computing resources now enable simulations with unprecedented
realism, allowing both one-dimensional simulations with a vast
amount of included physics and full three-dimensional simulations
albeit with less included physics~\citep{caldertownsley2018}.
In the area of stellar evolution, recent investigations revisiting
late-time evolution of roughly 8 \Msun~stars indicate that under
the right circumstances, ``hybrid" white dwarfs having a C-O core surrounded by O-Ne
mantle may form \citep{siess2009,denissenkovetal2013}. These hybrid
white dwarfs are thought to form when mixing at the lower convective boundary quenches
C burning in an asymptotic giant branch (AGB) star, leaving unburned C
in the core. The situation is at best uncertain, however, and the
results depend on assumptions about convective
overshoot that have been questioned {\color{blue} [questioned for what reasons?]} \citep{chenetal2014,lecoanetetal16,lattanzioetal2017}.

@@

In either {\color{red}ANY} case, the hybrid WD will experience a period of cooling, which will
make the core-mantle interface subject to convective
instability \citep{brooksetal2017,schwabgaraud2018}.
{\color{blue} [can we say a bit more about how this cooling-caused instability works?]}
In the case of the
hybrid WD accreting and approaching the Chandrasekhar mass, accretion will
heat the core and start C fusion, leading to a period of ``simmering" prior
to the explosion \citep{PiroBild08}. The upshot is that there is likely to be considerable
mixing after the hybrid forms that may homogenize the composition
\citep{denissenkovetal2015,brooksetal2017,schwabgaraud2018}.

@@

The process of mapping the \MESA\ models began by converting
the \MESA\ model to a uniform grid of $4$~km resolution.
{\color{blue}We mass-weighted and averaged quantities in \MESA\ zones with spacing less than $4$~km
and quadratically interpolated quantities in \MESA\ zones with spacing greater than $4$~km.}
At this point,
abundances of nuclides from the \MESA\ models were aggregated
into the abundances tracked by the flame capturing scheme in
\FLASH.
The most abundant isotopes in the model were \C{12}, \Ox{16},
and \Ne{20}, which are symmetric (the number of neutrons
equals the number of protons). Similarly to~\citet{willcoxetal2016},
other neutron-rich isotopes in the initial model were combined into
\Ne{22}, which serves as a proxy for metallicity. The aggregation
accounts for the $Y_e$ of the full set of nuclides, and \Ne{20}
and \Ox{16} were constrained to be in the same ratio in both
sets of abundances. \figref{fig:init_conds} shows the initial
profile of the one-dimensional models.

The original \MESA\ model was in equilibrium, but to ensure
the uniformly-gridded model was in hydrostatic equilibrium we
constructed the appropriate pressure profile. This was done by
integrating for the pressure in each zone from the central point up,
accounting for the local acceleration of gravity, temperature,
composition, and the mass below and enclosed by the zone. As
with \citet{willcoxetal2016} we used the EOS routine from
\CASTRO\ ~\citep{timmes.swesty:accuracy,castro1}.\footnote{The
Helmholtz EOS table used in \CASTRO\ is in the public BoxLib
Microphysics repository at
  \url{https://github.com/BoxLib-Codes/Microphysics.git} (commit
  hash $\mathrm{45ed859b6c1dc80d831d93f9728986d6ad6e1ddc}$).} As
 with the models in \citet{willcoxetal2016}, this procedure produced
 a structure that was stable in \FLASH, with fluctuations in central
density less than $3\%$, for at least $5$ seconds with no energy
deposition.

@@

The simulations performed for this study consisted of a suite of
30 two-dimensional simulations of
thermonuclear supernova
explosions from hybrid C-O-Ne progenitors in the DDT explosion paradigm. These
were compared to a suite of supernova simulations from traditional
C-O progenitors. In both suites, the simulation
begins with a progenitor model mapped to the two-dimensional
\FLASH\ grid. The burning is initiated with a ``match head,'' a region
in the white dwarf's center that is fully burned to nuclear statistical
equilibrium (NSE).
This initially burned region ignites a deflagration, a subsonic
flame, and because the match head was perturbed it is unstable to
the Rayleigh-Taylor instability with the result that buoyant plumes
rise. The star is partially consumed during this deflagration phase,
and the star responds by expanding.
When a plume reaches a specified density,
a detonation is initiated, and the simulation continues until
the expanding star reaches low densities at which point burning
effectively ends. The paragraphs in this section provide details
of the implementation of this method.

The C-O and C-O-Ne models have the same central temperature
and density, $8.2\times 10^8$~K and $2.2\times 10^9$~g~cm$^{-3}$ respectively.
The C-O model's composition consisted of \C{12}, \O{16}, and \Ne{22} in mass fractions 0.4, 0.57, and 0.02 respectively in the convective core and 0.5, 0.48, and 0.02 outside the convective core.
The C-O-Ne model used in flash had composition consisting of \C{12}, \O{16}, \Ne{20}, and \Ne{22}.
The mass fractions in the central convection zone were 0.14, 0.68, 0.23, and 0.05 respectively, and vary smoothly outside the convection zone as shown in Figure~\ref{fig:init_conds}.

@@

The expected final production of \Ni{56}, after NSE freeze-out, as a function of simulation time
for 10 simulations of explosions from each progenitor is shown in Figure \ref{fig:nithybrid}.
On average, the C-O model reaches the DDT phase sooner then the
hybrid model.
A few simulations ended early, but after reaching their final yields, due to problematic physical conditions in the fluff.
{\color{red}TODO: confirm.}

\begin{figure}
\includegraphics[width=\columnwidth]{figures/ige_v_mass_at_high_dens.pdf}
\caption{\label{fig:masshighdens}
Final IGE yield
vs.\ mass above $2\times 10^7~\grampercc$
at the time of the first DDT in a simulation.  Individual
realizations are shown (C-O-Ne
progenitors in blue, C-O in red) along with rectangles of length
$\pm 1 \sigma$ along each axis and centered at their average values
for C-O and C-O-Ne.  Similar rectangles found in W16 for a progenitor without
mixing during cooling are shown in a lighter shade with dashed borders.
}
\end{figure}

@@

We were also able to compare the results from these mixed
hybrid progenitors to those of the earlier unmixed progenitors.
\citep{willcoxetal2016} reported that compared to the unmixed
hybrid progenitors, C-O models consistently yielded a greater IGE mass. 
They interpreted this result as following from 
the lower \C{12} abundance in the hybrid models and the fact that \Ne{20}-rich fuel burns to cooler 
temperatures than the fuel in traditional C-O models, which. results in slower burning and thus a lower
overall IGE yield. Our yields from explosions from the mixed progenitors did not
show the same diminution in IGE yield. We attribute this difference to the higher
central density required for ignition of these mixed hybrid models.
The results of explosions from the mixed hybrid models 
produced a smaller fraction of IGEs in the form of \Ni{56} than the unmixed hybrid case, 
and we attribute this difference 
to both the central density and the location, center or off-center, of deflagration ignition,
which determine the amount of electron capture during burning and hence the relative abundance
of \Ni{56} to IGEs.

>>>>

Thermonuclear (Type Ia) supernovae (\SNeIa) are bright stellar explosions
thought to occur when approximately 1.0 \Msun\ of material composed principally
of C and O burns under degenerate conditions. This class of supernovae is
known to synthesize much of the Fe-group elements found in the galaxy, and
the light curves of these events have a special property that allows
standardization \blust{of their light curves and thus use}
as distance indicators for cosmological studies~\citep{phillips:absolute}.
This use resulted in the discovery of the acceleration of the expansion of
the Universe and thus the inference of Dark
Energy~\citep{riess.filippenko.ea:observational,
perlmutter.aldering.ea:measurements,leibundgut2001}, and these events
remain critical distance indicators for cosmological studies~\citep{weinbergetal2013}.
The special property of the light curve is thought to follow
from the fact that the source of luminosity, the radioactive decay
of \Ni{56} synthesized by the thermonuclear burning, is also the
principal source of opacity~\citep{Pinto2001The-type-Ia-sup}, giving
the Phillips relation between the brightness of an event and the
rate of decline of the B-band magnitude from maximum~\citep{phillips:absolute}.

@@

The single degenerate picture posits a white dwarf gaining mass
from a main-sequence companion. The process relies on a long
period of accretion combined with either steady burning or a
series of nova explosions. During this accretion phase, a traditional
C-O WD must gain at least
$\sim 0.4$ \Msun\ as needed for it to approach the
limiting Chandrasekhar mass~\citep{starrfieldetal2012}. As it approaches
the Chandrasekhar limit, conditions in the compressed core are right
to ignite the thermonuclear burning that will incinerate the star.

@@

A particular problem was that most models included a massive He layer on
the white dwarf. Such a configuration leads to the result that
intermediate and heavy elements synthesized in the He detonation appear in
the outer parts of the ejecta, which does not match observation~\citep{HoefKhok96,
hoeflichetal96,finkhillebrandtroepke2007,simetal2010}.  \citet{bildstenetal2007},
however, found that fairly thin He layers could flash on sub-Chandrasekhar
mass white dwarfs, partially resolving the issue and encouraging
further research~\citep{simetal2012,brooksetal2015, shenetal2018,
glasneretal2018}.

@@

Subsequent work allayed this concern {\color{blue} by demonstrating
that a low accretion rate from the disrupted secondary, $ < 3 \times 
10^{-6}$ \Msun per year, did not heat the primary enough for ignition}
~\citep{kawai1987,saionomoto2004,yoonetal2007,lorenaguilaretal2009, 
Shenetal12, pakmoretal2012b}
and active research continues, motivated by the need to explain bright events
and the estimated population of WD pairs. Contemporary
research focuses on variations on the merger idea, including inspiraling pairs,
collisions, violent mergers, and the ``core-degenerate'' model in which the merger
takes place in a common envelope~\citep{raskinetal2009,pakmoretal2011,kashi:2011,
pakmoretal2012a,Shenetal12,katzetal2016}.

@@

Modern computing resources now enable simulations with unprecedented
realism, allowing both one-dimensional simulations with a vast
amount of included physics and full three-dimensional simulations
albeit with less included physics~\citep[][and references therein]{caldertownsley2018}.
In the area of stellar evolution, recent investigations revisiting
late-time evolution of roughly 8 \Msun~stars indicate that under
the right circumstances, ``hybrid" white dwarfs having a C-O core surrounded by O-Ne
mantle may form \citep{siess2009,denissenkovetal2013}. These hybrid
white dwarfs are thought to form when mixing at the lower convective boundary quenches
C burning in an asymptotic giant branch (AGB) star, leaving unburned C
in the core. The situation is at best uncertain, however, and the
results depend on assumptions about convective
overshoot that have been questioned~\citep{chenetal2014,lecoanetetal16,lattanzioetal2017}.

@@

In either {\color{red}ANY} case, the hybrid WD will experience a period of cooling, 
which will reduce the stabilizing temperature gradient (the O-Ne layer is hotter 
than the C-O core) and allow the unstable composition gradient to drive 
thermocompositional convection~\citep{brooksetal2017,schwabgaraud2019}.
In the case of the
hybrid WD accreting and approaching the Chandrasekhar mass, accretion will
heat the core and start C fusion, leading to a period of ``simmering" prior
to the explosion, also mixing the interior \citep{PiroBild08}. The upshot is 
that there is likely to be considerable mixing after the hybrid forms that will 
homogenize the composition \citep{denissenkovetal2015,brooksetal2017,schwabgaraud2019}.

@@

The process of mapping the \MESA\ models began by converting
the \MESA\ model to a uniform grid of $4$~km resolution.
{\color{blue}To do this, \MESA\ quantities were mass-weighted and averaged in zones with spacing less than $4$~km
and quadratically interpolated in zones with spacing greater than $4$~km.}
At this point,
abundances of nuclides from the \MESA\ models were aggregated
into the abundances tracked by the flame capturing scheme in
\FLASH.
The most abundant isotopes in the model were \C{12}, \Ox{16},
and \Ne{20}, which are symmetric (the number of neutrons
equals the number of protons). Similarly to~\citet{willcoxetal2016},
other neutron-rich isotopes in the initial model were combined into
\Ne{22}, which serves as a proxy for metallicity. The aggregation
accounts for the $Y_e$ of the full set of nuclides, and \Ne{20}
and \Ox{16} were constrained to be in the same ratio in both
sets of abundances. \figref{fig:init_conds} shows the initial
profile of the one-dimensional models.

The original \MESA\ model was in equilibrium, but to ensure
the uniformly-gridded model was in hydrostatic equilibrium we
constructed the appropriate pressure profile. This was done by
integrating for the pressure in each zone from the central point up,
accounting for the local acceleration of gravity, temperature,
composition, and the mass below and enclosed by the zone. As
with \citet{willcoxetal2016}, we used the EOS routine from
\CASTRO\ ~\citep{timmes.swesty:accuracy,castro1}.\footnote{The
Helmholtz EOS table used in \CASTRO\ is in the public BoxLib
Microphysics repository at
  \url{https://github.com/BoxLib-Codes/Microphysics.git} (commit
  hash $\mathrm{45ed859b6c1dc80d831d93f9728986d6ad6e1ddc}$).} 
This procedure produced a structure that was stable in \FLASH, with 
fluctuations in central density less than $3\%$, for at least $5$ 
seconds with no energy deposition.

@@

The simulations performed for this study consisted of a suite of
30 two-dimensional simulations of
thermonuclear supernova
explosions from hybrid C-O-Ne progenitors in the DDT explosion paradigm. These
were compared to a suite of supernova simulations from traditional
C-O progenitors. In both suites, the simulation
begins with a progenitor model mapped to the two-dimensional
\FLASH\ grid. The burning is initiated with a ``match head,'' a region
in the white dwarf's center that is fully burned to nuclear statistical
equilibrium (NSE).
This initially burned region ignites a deflagration, a subsonic
flame, and because the match head was perturbed it is unstable to
the Rayleigh-Taylor instability with the result that buoyant plumes
rise. The star is partially consumed during this deflagration phase,
and the star responds by expanding.
When a plume reaches a specified density,
a detonation is initiated, and the simulation continues until
the expanding star reaches low densities at which point burning
effectively ends. This section provide details
of the implementation of this method.

The C-O and C-O-Ne models have the same central temperature
and density, $8.2\times 10^8$~K and $2.2\times 10^9$~g~cm$^{-3}$ respectively.
The C-O model's composition consisted of \C{12}, \Ox{16}, and \Ne{22} in mass fractions 0.4, 0.57, and 0.02 respectively in the convective core and 0.5, 0.48, and 0.02 outside the convective core.
The C-O-Ne model used in flash had composition consisting of \C{12}, \Ox{16}, \Ne{20}, and \Ne{22}.
The mass fractions in the central convection zone were 0.14, 0.68, 0.23, and 0.05 respectively, and vary smoothly outside the convection zone as shown in Figure~\ref{fig:init_conds}.

@@

The expected final production of \Ni{56}, after NSE freeze-out, as a function of simulation time
for 10 simulations of explosions from each progenitor is shown in Figure \ref{fig:nithybrid}.
On average, the C-O model reaches the DDT phase sooner then the
hybrid model.
A few simulations ended early, but after reaching their final yields. We attribute
these occurences as being due to problematic physical conditions in the ``fluff," the
low-density gas on the grid outside of the star. {\color{red}TODO: confirm if
possible. If not, leave as is, which is speculative.}

\begin{figure}
\includegraphics[width=\columnwidth]{figures/ige_v_mass_at_high_dens.pdf}
\caption{\label{fig:masshighdens}
Final IGE yield
vs.\ mass above $2\times 10^7~\grampercc$
at the time of the first DDT in a simulation.  Individual
realizations are shown (C-O-Ne
progenitors in blue, C-O in red) along with rectangles of length
$\pm 1 \sigma$ along each axis and centered at their average values
for C-O and C-O-Ne.  Similar rectangles found in \citet{willcoxetal2016} for a progenitor without
mixing during cooling are shown in a lighter shade with dashed borders.
}
\end{figure}

@@

Comparing the relative abundances of \Ni{56} to IGE between the
mixed hybrid progenitors and traditional C-O models indicated
substantial differences, with the hybrid progenitors producing
a lower ratio of \Ni{56} to IGEs. This difference was particularly
pronounced during the deflagration phase of the explosion, and 
from that we conclude that the evolution of high-density material
and the amount of neutronization during the burning at high densities
is critical to the evolution and outcome of the explosion.

We were also able to compare the results from these mixed
hybrid progenitors to those of the earlier unmixed progenitors.
\citet{willcoxetal2016} reported that compared to the unmixed
hybrid progenitors, C-O models consistently yielded a greater IGE mass. 
They interpreted this result as following from 
the lower \C{12} abundance in the hybrid models and the fact that \Ne{20}-rich fuel burns to cooler 
temperatures than the fuel in traditional C-O models, which. results in slower burning and thus a lower
overall IGE yield. Our yields from explosions from the mixed progenitors did not
show the same diminution in IGE yield. We attribute this difference to the higher
central density required for ignition of these mixed hybrid models.
The results of explosions from the mixed hybrid models 
produced a smaller fraction of IGEs in the form of \Ni{56} than the unmixed hybrid case, 
and we attribute this difference 
to both the central density and the location, center or off-center, of deflagration ignition,
which determine the amount of electron capture during burning and hence the relative abundance
of \Ni{56} to IGEs.
